No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-11-14 paper/GEB Submission/manuscript_GEB.Rmd
Small changes, fix table label

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: apa-6th-edition.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  # rticles::arxiv_article:
  #   keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - time series

@@

We also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait. Wholesale shifts in the `r colorize("observed", "blue")` trait space due to changes of trait means could occur even while the shape `r colorize("and dispersion", "blue")` of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained `r colorize("through time", "blue")`. CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

change_group_ses_desc %>%
  select(type = change_description, everything()) %>%
  mutate(n = as.character(n),
           n = ifelse(n == "7", "32", n)) %>%
  mutate(type = ifelse(n == "32", "no change", type)) %>%
  rename(`Study Count` = `n`) %>%
     kable(format = "latex",
        caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness. Trends direction is indicated by a positive or negative sign, with a 0 indicating no trend.", align = "c",
        latex_options= c("scale_down", "hold_position")) %>% 
  collapse_rows(columns = c(1,2)) %>%
  column_spec(1, width = "4cm") #%>%
  #save_kable(here::here("figures/tables/study_change_table.jpeg"))

>>>>

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: apa-6th-edition.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
   rticles::arxiv_article:
     keep_tex: true
  #word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - time series

@@

To assess compositional changes, we also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait. Wholesale shifts in the `r colorize("observed", "blue")` trait space due to changes of trait means could occur even while the shape `r colorize("and dispersion", "blue")` of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained `r colorize("through time", "blue")`. CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` Fore each linear mixed effects model we estimated within-level slopes and calculated between-level contrasts for each categorical variable using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

change_group_ses_desc %>%
  select(type = change_description, everything()) %>%
  mutate(n = as.character(n),
           n = ifelse(n == "7", "32", n)) %>%
  mutate(type = ifelse(n == "32", "no change", type),
         type = ifelse(type == "functional change", "functional diversity change", type)) %>%
  rename(`Study Count` = `n`) %>%
     kable(format = "latex",
        caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness. Trends direction is indicated by a positive or negative sign, with a 0 indicating no trend.", align = "c",
        latex_options= c("scale_down", "hold_position")) %>% 
  collapse_rows(columns = c(1,2)) %>%
  column_spec(1, width = "4cm") #%>%
  #save_kable(here::here("figures/tables/study_change_table.jpeg"))
